Supporting Information

### 2015-08-27 ms.Rmd
code for re-ordering mfuzz plot. need to run clustering and hypergeom again to verify

At least `r round(min(st_libStats$"Reads (M)"), 0)` million 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after removal of reads likely to have originated from rRNA and tRNA. Using cutoffs calculated from the 95^th^ percentile of distributions of expression values estimated from intergenic regions for each library (*i.e*. reads that were not generated from genes), `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries. As a preliminary quality check, the qualitative expression patterns of a selection of genes were compared with patterns that had previously been reported by *in situ* hybridisation (currently `r I(pasteLabel("Figure ", figCount, "f_reviewInSitu"))`, **move figure to supporting table**).

@@

To recover common expression patterns in genes detected in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do]. Expressed genes were ranked by variance, and standardised, VST-transformed read counts for the 25% most-variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo]. To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figures S", s_figCount, "sf_mfuzzCentroids"))` `r I(pasteLabel(" and S", s_figCount, "sf_mfuzzPca"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`). Clusters were recovered that contain genes that gradually decrease or increase in expression during development (clusters **x**, **x**, **x** and **x**), or that are apparently turned on or off in primary and secondary branch meristems (clusters **x** and **x** respectively) along with genes with more complex expression patterns (clusters **x** and **x**).

@@

Examination of the clustered genes suggested that transcription factors were overrepresented in most clusters (**annotated cluster tables in SI**). To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests. Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database (**reference**), a hypergeometric test was carried out if there were at least two members of one family in a given cluster. After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`). For example, members of the ABI3VP1 family were overrepresented in clusters **x** and **x**, both of which contain genes with higher transformed read counts in the rachis meristem than in samples from later stages of development. The MADS family was overrepresented in clusters **x** and **x**, suggesting that they could be involved in both early and late stages of inflorescence development.

@@

**cluster y-axis of GSEA plot w/in protein type**

These patterns are not intended to represent each member of each gene family. For example, there are several MADS genes that have lower expression in the SM than in the RM, but expressed MADS genes were detected at a higher level overall in SM. Despite this caveat, there are some notable modes of expression shared between different families that may indicate roles in inflorescence development. In the GSEA, the most pronounced change in expression appears to occur between the RM and branch meristem samples. Several families (including TUBBY-like, GROWTH-REGULATING FACTOR, bZIP and ABI3VP1 TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types. Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enrinched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.

@@

>>>>

At least `r round(min(st_libStats$"Reads (M)"), 0)` million 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million reads that mapped to genes after excluding reads likely to have originated from rRNA and tRNA. Using strict cutoffs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in at least one of the libraries. As a preliminary quality check, the qualitative expression patterns of a selection of genes were compared with patterns that had previously been reported by *in situ* hybridisation (currently `r I(pasteLabel("Figure ", figCount, "f_reviewInSitu"))`, **move figure to supporting table**).

@@

To recover common expression patterns in genes detected in the meristem samples, read counts were transformed using the variance-stabilizing transformation (VST) included in the `DESeq2` package [@Love:2014do]. Expressed genes were ranked by variance, and standardised, VST-transformed read counts for the 25% most-variable genes were clustered using the fuzzy *c*-means algorithm implemented in the `Mfuzz` package [@Futschik:2005vo]. To determine the number of cluster cores (*c*), the minimum distance between cluster centroids, the formation of empty clusters, and PCA plots of cluster members were monitored for clusters produced with *c* values between 2 and `r maxClust` (`r I(pasteLabel("Figures S", s_figCount, "sf_mfuzzCentroids"))` `r I(pasteLabel(" and S", s_figCount, "sf_mfuzzPca"))`), leading to the use of `r as.character(english::english(c))` cluster cores (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`). There are several expression patterns descibed by the clusters: genes that increase or decrease steadily during development (clusters `r getClusterName('inc', 'inc')` and `r getClusterName('dec', 'dec')`), genes that change expression between RM and PBM and remain constant from PBM to SM (clusters `r getClusterName('inc', 'small')` and `r getClusterName('dec', 'small')`), genes that are constant between RM and PBM and then change between PBM and SM (clusters `r getClusterName('small', 'inc')` and `r getClusterName('small', 'dec')`). There were also two weaker clusters containing fewer genes, which had complex expression patterns involving changes in expression particular to the PBM (clusters `r getClusterName('inc', 'dec')` and `r getClusterName('dec', 'inc')`).

@@

Examination of the clustered genes suggested that transcription factors were overrepresented in most clusters (**annotated cluster tables in SI**). To determine whether particular TF families were associated with particular patterns of expression, the enrichment of TF family members in each cluster was calculated using hypergeometric tests. Using the list of transcription factor and other transcriptional regulator families available on the Plant Transcription Factor Database (**reference**), a hypergeometric test was carried out if there were at least two members of one family in a given cluster. After adjustment for multiple comparisons, each cluster had a significant enrichment of at least one family (`r I(pasteLabel("Table ", tableCount, "t_hypergeom"))`). For example, members of the ABI3VP1 family were overrepresented in clusters **x** and **x**, both of which contain genes with higher transformed read counts in the rachis meristem than in samples from later stages of development. The MADS family was overrepresented in clusters **x** and **x**, suggesting that they could be involved in both early and late stages of inflorescence development. **run clustering script again and check these statements!!!***

@@

These patterns are not intended to represent each member of each gene family. For example, there are several MADS genes that have lower expression in the SM than in the RM, but expressed MADS genes were detected at a higher level overall in SM. Despite this caveat, there are some notable modes of expression shared between different families that may indicate roles in inflorescence development. In the GSEA, the most pronounced change in expression appears to occur between the RM and branch meristem samples. Several families (including TUBBY-like, GROWTH-REGULATING FACTOR, bZIP and ABI3VP1 TF genes and Aux/IAA genes) are strongly expressed in RM and lower in the other meristem types. Another group, including mTERF, MYB, SBP and FHA TF genes and SET, PHD and Jumonji regulator genes, is enriched in one or both of the PBM and ePBM/SBM samples but depleted or not enriched in the RM samples.

@@

## Plant material and sampling

*Oryza sativa* ssp. japonica cv. Nipponbare plants were grown in a growth chamber at 70% relative humidity at 30°C during the day and 26°C at night. After 7 weeks in long day conditions (LD; 14 hours light and 10 hours darkness) plants were transferred to short day conditions (SD; 10 hours light and 14 hours darkness) to induce floral transition. Inflorescence meristems were harvested from 6 to 14 days after the change of photoperiod.

## Tissue embedding, LASER microdissection and sequencing

Each inflorescence was harvested in 2 mL ice-cold 3:1 ethanol:acetic acid fixative solution, infiltrated (=**filtered**?) twice under a mild vacuum for 15 minutes and stored for 20 hours in fresh fixative at 4°C. Embedding for dissection was performed as described by @Mantegazza:2014di. Tissues were cut into 8 µm sections on an RM2155 or RM2255 microtome (Leica) and dissected on a LMD6000 or LMD7000 LASER dissector (Leica). RNA was amplified using the ARCTURUS PicoPure RNA Isolation Kit (ThermoFisher) and assayed on a 2100 Bioanalyzer with the RNA 6000 Pico Kit (Agilent) before amplification and cDNA synthesis with the Ovation RNA-Seq System V2 (NuGEN). Library preparation with **which kit** and 50-base single-end sequencing on the HiSeq 2000 platform (Illumina) were performed by IGA Technology Services (Udine, Italy).

## Data analyses

Analysis of the sequenced libraries was carried out with custom `bash` and `R` (**ref**) scripts, which were arranged into a pipeline using the `ruffus` package for `python3` (**ref**). All of the code used for the analysis and to generate text, figures and tables for this report, along with the versions and paramaters of the software used, are available with revision history in a public GitHub repository at https://github.com/evoreprice/lmdPaper.

Briefly, adaptor trimming was performed with `cutadapt` (**ref**) and reads were mapped with the MSUv7 genome and annotation (**ref**) downloaded from Phytozome 10.3 (**ref**) using `rna-star` (**ref**) in 2-pass mode. The `quantMode` argument of `rna-star` was used for read counting. rRNA and tRNA contamination was estimated using `htseq-count` (**ref**) to count the number of reads that mapped to regions annotated as tRNA or rRNA on the Rap-DB (**ref**) or to regions where reads simulated with `wgsim` (**ref**) from rRNA sequences in the Plant Repeats Database (**ref**) also mapped. Differential expression analysis to calculate Log~2~ fold change (L~2~FC) values and transformed read counts was performed with the `DESeq2` package (**ref**). To determine a strict cutoff for unexpressed genes, expression values in transcripts per million (TPM) were calculated for each gene (**2 refs**). These values were compared to pseudo-expression values calculated from intergenic regions of the genome, using intervals of a similar size distrubution to those used for calculating the 'genic' expression values. For each library, the cutoff was placed at the 95^th^ percentile of the distribution of intergenic expression values. Transformed counts and L~2~FC values from expressed genes were used for soft clustering with the `Mfuzz` package (**ref**) and geneset enrichment analysis with the `gage` package (**ref**) respectively. All plots were produced with the `ggplot2` package (**ref**).

### 2015-08-27 si.Rmd
code for re-ordering mfuzz plot. need to run clustering and hypergeom again to verify

knitr::opts_chunk$set(fig.path='figures/',
               cache.path='cache/',
               dev=c("png", "cairo_pdf"),
               fig.width=6.614,
               fig.height=6.614,
               fig.align = 'center',
               dpi=300,
               fig.show='hold',
               cache=FALSE,
               echo=FALSE,
               results="hide",
               message=FALSE,
               warning=FALSE)

>>>>

knitr::opts_chunk$set(fig.path='figures/',
               cache.path='cache/',
               dev=c("png", "cairo_pdf"),
               fig.align = 'center',
               dpi=300,
               fig.show='hold',
               cache=FALSE,
               echo=FALSE,
               results="hide",
               message=FALSE,
               warning=FALSE)
